---
name: Preferences & Utility
---

## Where Does Demand Come From?

Imagine the market for coffee on campus

- Additionally, picture that the only other good that can be purchased is tea

- On a basic level, [demand for coffee]{.hii} is derived from individual's choosing how to divide their income between coffee and tea

```{r, fig.align='center'}
library(ggplot2)

# Create data for the supply and demand curves
quantity <- c(10, 20, 30, 40, 50, 60)
price_supply <- c(1, 2, 3, 4, 5, 6)
price_demand <- c(6, 5, 4, 3, 2, 1)

# Create a data frame
curve_data <- data.frame(quantity, price_supply, price_demand)

# Create the plot
ggplot(curve_data, aes(x = quantity)) +
  geom_line(aes(y = price_supply), color = "blue", size = 1, linetype = "solid") +
  geom_line(aes(y = price_demand), color = "red", size = 1, linetype = "solid") +
  labs(
    y = "Price",
    x = "Quantity"
  ) +
  theme_minimal() +
  annotate("text", x = 60, y = 5.6, label = "S", color = "blue", size = 8) +
  annotate("text", x = 60, y = 1.3, label = "D", color = "red", size = 8)
```

---

## Where Does Demand Come From?

This is the problem we will be dealing with through the first half of the course

> How does an individual allocate a [finite amount of resources]{.hii} between [two goods]{.hii}?

. . .

<br>

[Note:]{.h .note} There are clearly more than just two goods out there, so how can this be useful?

[The main critique I always hear about economics courses is that they're unrealistic. That's mostly true, but we can learn about the aggregate by simplifying and making assumptions]{.h .note}
---

## Where Does Demand Come From?

- We will frame the decisions as a two-good model where you may choose between:

  - Food and durable consumption
  - Leisure (not working) and consumption (paid for by earning a wage)
  - Consumption now and consumption later

> [The key takeaway here is that we can frame many important choices as "two-good" decisions]{.h .note}

- This makes things simpler for us to solve while still maintaining some sense of the real world

---

## Determinants of Consumption

Consumption of any single good has two parts:

- How it [BENEFITS]{.hii} the consumer 

  - We call this [UTILITY]{.hii}

- What it [COSTS]{.hii} the consumer

  - What we give up to purchase the good

. . .

Let's see what this means through a 1-good example $\rightarrow$ Beer

- Imagine the following scenario:

  - You just arrived at the bar and have had zero drinks so far
  - Each beer costs the same: $4

::: {.notes}
Beer is a great example of a good with complicated benefits
:::

---

## Beer {.smaller}

| Number of Beers | Overall Level of Happiness | Change in Level of Happiness |
|:---------------:|:---------------------------:|:-----------------------------:|
|        0        |             0               |               -               |
|        1        |             10              |               10              |
|        2        |             25              |               15              |
|        3        |             35              |               10              |
|        4        |             40              |               5               |
|        5        |             42              |               3               |
|        6        |             30              |              -12              |

. . .

:::: {.columns}

:::: {.column width="50%"}
If beer were free, how many beers should this person drink?

::: {.fragment }

[$$5$$]{.hi}

:::

::::

:::: {.column width="50%"}

::: {.fragment}

Now recall that beer costs $4, how many beers should this person drink?

::: {.fragment}

[$$4$$]{.hi}

:::

:::

::::

:::: 

---

## Intuition Behind "Choice" in Economics

You [cannot]{.hi-red} simply find the consumption amount that makes you the happiest. 
[But why?]{.hi}

. . .

> The goal is to maximize your [utility]{.hii} whilst acknowledging you have [constraints]{.hii}
>
> The choice is simple: [consume an additional unit until the cost of doing so outweighs the benefit]{.hii}

<br>

. . .

[Commit this idea to memory: it is the crux of economics and drives everything we will be doing]{.hi .note}

[We maximize utility up to the point that it does not make sense to do so]{.hi .note}

---

## Back to Beer {.smaller}

| Number of Beers | Overall Level of Happiness | Change in Level of Happiness |
|:---------------:|:---------------------------:|:-----------------------------:|
|        0        |             0               |               -               |
|        1        |             10              |               10              |
|        2        |             25              |               15              |
|        3        |             35              |               10              |
|        4        |             40              |               5               |
|        5        |             42              |               3               |
|        6        |             30              |              -12              |

Some Questions:

:::: {.columns}

::: {.column width="50%"}
- What is the marginal benefit:
  - When you have not consumed any beer?
  - When you have already consumed 3 beers?
  - When you have already consumed 5 beers?
:::

::: {.column width="50%"}
- What is the marginal cost of beer? 
  - Does it change as we consume more?
:::

::::

---

## Being and Thinking at the Margin

We found two important values:

:::: {.columns}

::: {.column width="50%"}
- [Marginal Benefit (MB)]{.hi}

::: {.fragment}
The additional benefit gained for an added unit of consumption
:::

:::

::: {.column width="50%"}
- [Marginal Cost (MC)]{.hi}

::: {.fragment}
The additional cost paid for an added unit of consumption 
:::

:::

::::

. . .

We can describe the decision-making process in a more formal manner:

- Initially: [MB]{.hi-green} > [MC]{.hi-red} $\rightarrow$ Consume more!
- Eventually: [MB]{.hi-green} < [MC]{.hi-red} $\rightarrow$ We went too far!
- At some point: [MB]{.hi-green} = [MC]{.hi-red} $\rightarrow$ Just right!

> [Ask yourself: Why must they be equal?]{.hi .note}

---

## What's Going On? Graphically 

```{r beer-utility, fig.align='center'}
# Load the ggplot2 library if not already loaded
library(ggplot2)

# Create a data frame with x values
data <- data.frame(x = c(0,1,2,3,4,5,6), y = c(0,10,25,35,40,42,30))

# Create the ggplot object
utility <- ggplot(data, aes(x = x, y = y)) +
  geom_line(color = "black", size = 2) +
  xlim(0,6) +
  labs(
    x = "Beers",
    y = "Utility"
  ) +
  theme_minimal()

utility
```

. . .

What matters for choice is the [marginal benefit]{.hi} of an additional beer

In other words, what matters is the change in [utility]{.hii} that occurs as we move to the right on the graph

. . .

> [Recall: A change in $y$ as $x$ increases is the derivative]{.h .note}

---

## What's Going On? Graphically - Derivatives

```{r beer-utility-derivative, fig.align='center'}
utility
```

The derivative of this function is the [marginal benefit]{.hi} of beer

[This is why derivatives are important]{.h .note}

We can use derivatives to figure out the optimal amount to consume

---

## What's Going On? Graphically - MB & MC

```{r beer-utility-derivative1, fig.align='center'}
utility
```

> [Recall:]{.hi .note} The optimal choice is the point where [MB]{.hi-green} = [MC]{.hi-red} 

> The [MC]{.hi-red} = 4. So we would choose the quantity of beers where [MB]{.hi-green} = 4 

---

## What's Going On? Mathematically

The utility function of beer we've been using is:

$$U(x) = -x^{2} + 12x$$

[Let's practice:]{.h} 
What would the optimal consumption amount be if the cost of beer is $2?

---

## A Two-Good Problem

The beer example was fairly straightforward. But we will be dealing with making choices between two goods. 

Before we dive in, a couple of things to consider are:

. . .

- When we spend our resources on one good, it cannot be spent on something else
  - Because we are making decisions amongst both things, we need a [cost relationship]{.hi} between them

. . .

- We do not measure our happiness (utility) in dollars
  - We need to find a way to choose without directly comparing costs and benefits across goods

--- 

## Enter Utility Functions {.smaller}

A [Utility Function]{.hi} is a function of two variables:

$$U = f(x,y)$$

[Some facts:]{.h}

- $x$ and $y$ are the (positive) amounts of goods you consume
- The function converts $x$ and $y$ to happiness (utility) from consuming the two goods

. . . 

[For Example:]{.h}

- Let's say that the utility I receive from Beer (B) and Soda (S) can be modeled as:

$$U = f(B,S) = 10B + 2S$$

- Before, we were only interested in the slope of Beer
- Now, we are interested in the [slopes]{.hii} of both Beer & Soda

---

## Utility Functions 

