---
name: calculus
---

## Starting Simple

```{r}
library(ggplot2)

# Create data for the line
x <- c(0, 2, 4, 6, 8)
y <- x * (1/2) + 5

line_data <- data.frame(x, y)

# Create the line plot
line_plot <- ggplot(line_data, aes(x, y)) +
  geom_line(color= "#b45c20", lwd = 2) +
  labs(x = "X", y = "Y") +
  theme_minimal()

print(line_plot)
```

We call the slope the [RISE]{.hi} over [RUN]{.hi} 

---

## Functions & Slopes

We can write that previous line as a mathematical function:

$$f(x) =  \dfrac{x}{2} + 5$$

Importantly, the slope for this function is [$1/2$]{.hi} [everywhere]{.hii}

. . .

- Everywhere means for [all values of $x$]{.hii}

- So no matter at what point of the line we calculate the slope, we will always find it to be [$1/2$]{.hi}

---

## Functions & Slopes

Slopes can take positive, negative, or zero values

And each of these let's us know how the function is behaving as we increse in $x$ (as we move to the right):

- Positive values mean the function is increasing

- Negative values mean the function is decreasing

- Zero values mean the function is staying constant (neither increasing or decreasing)

---

```{r}
library(ggplot2)

# Create x values
x <- seq(0, 5, by = 0.5)

# Calculate y values for each line
y_positive_slope <- 0.5 * x + 2
y_negative_slope <- -0.5 * x + 5
y_zero_slope <- rep(3, length(x))

# Create a data frame
line_data <- data.frame(x, y_positive_slope, y_negative_slope, y_zero_slope)

# Create the line plot
line_plot <- ggplot(line_data, aes(x)) +
  geom_line(aes(y = y_positive_slope), color = "blue", linetype = "solid", size = 1) +
  geom_line(aes(y = y_negative_slope), color = "red", linetype = "solid", size = 1) +
  geom_line(aes(y = y_zero_slope), color = "green", linetype = "solid", size = 1) +
  labs(x = "X", y = "Y") +
  theme_minimal()

print(line_plot)
```

---

## Functions & Slopes {.smaller}

The previous functions are called linear functions:

- Slopes are constant at all values of $x$
- They are behaving in an equal manner at all points

. . .

<br>

We will mostly be working with non-linear functions, which simply mean they have some form of curvature

- We will use quadratic and cubic functions

    - Quadratic: $f(x) = 3x^{2} + 5x + 10$

    - Cubic: $f(x) = x^{3} - 2x^{2} + x - 5$

. . .

What we truly care about is their slope. And there is a simpler way to find the slope of a function.