---
name: Why Econ
---

## Foundations

What we will be learning in this course will seem simplistic and reductive to the world we actually observe

Before we jump into Intermediate Microeconomics, we need to refresh our memories of more basic concepts from EC 201:

- Demand

- Supply 

- Economic Assumptions

# Basics Review {.inverse .note}

---

## Demand {.align-center}

```{r} 
library(ggplot2)

# Create data for demand and supply curves
price <- c(1, 2, 3, 4, 5)
quantity_demanded <- c(10, 8, 6, 4, 2)
quantity_supplied <- c(2, 4, 6, 8, 10)

df <- data.frame(price, quantity_demanded, quantity_supplied)

# Create demand and supply curve plot
demand_plot <- ggplot(df, aes(x = quantity_demanded, y = price)) +
  geom_line(color = "blue") +
  labs(x = "Quantity", y = "Price") +
  theme_minimal() 

print(demand_plot)
```

---

## Demand

- [Demand]{.hp} is the [QUANTITY]{.hi} of a good that consumers want at a given [PRICE]{.hi}

. . .

- The [Demand Curve]{.hp} shows the [RELATIONSHIP]{.hi} between [PRICES]{.hi} and [QUANTITY]{.hi} of a good for consumers

. . .

We will make the distinction between [Individual Demand]{.hii} and [Market Demand]{.hii}

[Individual Demand:]{.hii} Single Consumer (One Person)

[Market Demand:]{.hii} All potential consumers (Everyone participating in the market)

. . .

[Note: The first half of the course will be about demand and how we can find it for individuals and then for an entire market]{.note .hii}

---

## Demand

Some [Factors]{.hi} that can affect demand are:

- Number of Consumers

- Changes to Income

- Consumer Preferences (Tastes)

- The Price of Other Goods

    - Complements
    - Substitutes

---

## Supply {.align-center}

```{r} 
library(ggplot2)

# Create data for demand and supply curves
price <- c(1, 2, 3, 4, 5)
quantity_demanded <- c(10, 8, 6, 4, 2)
quantity_supplied <- c(2, 4, 6, 8, 10)

df <- data.frame(price, quantity_demanded, quantity_supplied)

# Create demand and supply curve plot
supply_plot <- ggplot(df, aes(x = quantity_supplied, y = price)) +
  geom_line(color = "red") +
  labs(x = "Quantity", y = "Price") +
  theme_minimal() 
print(supply_plot)
```

---

## Supply

- [Supply]{.hp} is the [QUANTITY]{.hi} of goods available for sale at a given [PRICE]{.hi}

. . .

- The [Supply Curve]{.hp} shows the [RELATIONSHIP]{.hi} between [PRICES]{.hi} and [QUANTITY]{.hi} of a good for producers

. . .

We also make the distinction between [Individual Supply]{.hii} and [Market Supply]{.hii}

[Individual Supply:]{.hii} A Single Producer (One Firm)

[Market Supply:]{.hii} All potential producers (All firms that are participating in the market)

. . .

[Note: The second half of the course will deal with how producers make their decisions]{.note .hii}

---

## Supply

Some [Factors]{.hi} that may affect supply are:

- Production Costs (Cost of Raw Materials, Technology, etc.)

- Number of Sellers (Number of firms in the market)

- Producers outside options

---

## Supply & Demand {.align-center}

```{r} 
library(ggplot2)

# Create data for demand and supply curves
price <- c(1, 2, 3, 4, 5)
quantity_demanded <- c(10, 8, 6, 4, 2)
quantity_supplied <- c(2, 4, 6, 8, 10)

df <- data.frame(price, quantity_demanded, quantity_supplied)

# Create demand and supply curve plot
supply_plot <- ggplot(df, aes(x = quantity_supplied, y = price)) +
  geom_line(color = "red") +
  geom_line(aes(x = quantity_demanded, y = price), color = "blue") + 
  labs(x = "Quantity", y = "Price") +
  theme_minimal() 

print(supply_plot)
```

[Put together they give us this classic graph]{.note .hii}

---

## Supply & Demand

Where these two are [equal to each other]{.hi-teal} (where they cross) we say it is the [market equilibrium]{.hi-teal}

The third part of the course will be about variations to the [classic market equilibrium]{.hii}

- What happens when the market is deep (has many producers) [(Perfect Competition)]{.hi}

- What happens when the market is shallow (only one producer) [(Monopoly)]{.hi}

- What about other forms of competition where firms have some market power but not total market control? [(Imperfect Competition)]{.hi}

---

## Models

We can use all these to make very complex models of what we think the world looks like

But to begin, we need to develop the basics of:

- What does [Demand]{.hi} look like?

- What does [Supply]{.hi} look like?

- What form of [Competition]{.hi} are we in?

> Once we put all of these together we are modeling a market. With a model, you can begin to analyze the effects of changes to [Model Inputs]{.hi-teal} on [Model Outputs]{.hi-bb}

---

## Assumptions

We will work with models you have seen before and some new ones that are classified as "classical economics"

. . .

These require certain assumptions in order to simplify the results:

- Assume that individuals are [Rational and Self-Interested]{.hii}

- Assume that markets have [No Externalities and Asymmetric Information]{.hii}

---

## Assumptions

These assumptions help make things simple and work to set a benchmark of sorts

They are often good approximations of the real world but are often not exact

We use these fundamental pieces of classical economics and build on top of them

<br>

[Note: I'm not attempting to describe the way the world works excatly, but rather give you a strating point for future endeavors]{.note .hi}

---

## Last Tool You Will Need

We will need calculus

It turns out that intermediate micro is fundamentally inseparable from multi-variate calculus

. . .

- I cannot show you where a demand curve comes from without using [(a tiny bit)]{.hi} of calculus

So the second thing we will do is discuss the math you will need in order to follow the material 